Inferring Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-07-02 doc/paper-lcr.tex
first draft of theory section

Previous attempts at reconstructing the line of sight mass distribution in
strong lens fields have focused on spectroscopic surveys to ascertain group
membership, and on understanding the external shear apparently required by the
strong lens models.  The probability density function (PDF) $\Pr(\xkappa)$ is
the target for any reconstruction analysis, and the key quantity for any
ensemble analysis, since by marginalising over it we can correctly propagate
the uncertainties due to weak lensing by line of sight mass structure. 
\flag{More needed here. Previous work on environments: Keeton, Zabludoff,
Auger, Momcheva, Wong, others.} Most recently, \citet{Wong++2011} ran Monte
Carlo reconstructions of the mass in nine lens fields, and compared the
resulting predicted shear distributions with the lens model external shear.
Their simulations show most of  the shear being generated by bright galaxies
within 2~arcminutes of the lens, and that both line of sight structures and
the group of galaxies in each lens plane contribute significant proportions of
the shear. The significant discrepancies between the lens model shear and the
Monte Carlo predictions are attributed to both the environment modelling and
the strong lens modelling itself.

@@

Time delay distances, and the mass sheet degeneracy - how kappa is degenerate in
strong lens models (and magnifications). Weak lensing by NFW halos. Accounting
for mass along the line of sight.  Approximations to full ray tracing: combining
kappas: simple sum, Keeton scaling, comments. Example ray traced kappa map from Stefan.

@@

\section{Halo Model}
\label{sec:model}

>>>>

Previous attempts at reconstructing the line of sight mass distribution in
strong lens fields have focused on spectroscopic surveys to ascertain group
membership, and on understanding the external shear apparently required by the
strong lens models.  The probability density function (PDF) $\Pr(\xkappa)$ is
the target for any reconstruction analysis, and the key quantity for any
ensemble analysis, since by marginalising over it we can correctly propagate
the uncertainties due to weak lensing by line of sight mass structure. 
\flag{More needed here. Previous work on environments: Keeton, Zabludoff,Auger, Momcheva, Wong, others.}
Most recently, \citet{Wong++2011} ran Monte
Carlo reconstructions of the mass in nine lens fields, and compared the
resulting predicted shear distributions with the lens model external shear.
Their simulations show most of  the shear being generated by bright galaxies
within 2~arcminutes of the lens, and that both line of sight structures and
the group of galaxies in each lens plane contribute significant proportions of
the shear. The significant discrepancies between the lens model shear and the
Monte Carlo predictions are attributed to both the environment modelling and
the strong lens modelling itself.

@@

%Time delay distances[✓], and the mass sheet degeneracy[✓] - how kappa is degenerate in
%strong lens models (and magnifications).

Strong lensing occurs when a massive object and background source are almost perfectly
aligned along a line of sight. The light from the background source is deflected by the
lens galaxy; this deflection allows multiple images of the background source to form
at stationary points of the time delay function. For a strong lens system in an otherwise
homogeneous Robertson--Walker universe, the excess time delay of an image at angular position
$\bmath{\theta}=(\theta_1,\theta_2)$ with unlensed source position $\bmath{\beta}=(\beta_1,\beta_2)$ is
\be \label{eq:T} 
\Delta t(\bmath{\theta},\bmath{\beta}) = \frac {1}{c} \frac{D_{\rm d} D_{\rm s}}{D_{\rm ds}} (1+z_{\rm d})\, \phi(\bmath{\theta},\bmath{\beta}),
\ee
where $z_{\rm d}$ is the redshift of the lens, $\phi(\bmath{\theta},\bmath{\beta})$ is
the Fermat potential, and $D_{\rm d}$, $D_{\rm s}$, and $D_{\rm ds}$
are, respectively, the angular diameter distance from observer to the lens, from observer
to the source, and from the lens to the source. The Fermat potential is
defined as
\be \label{eq:FP}
\phi(\bmath{\theta},\bmath{\beta})\equiv \left[\frac{(\bmath{\theta}-\bmath{\beta})^2}{2}-\psi(\bmath{\theta}) \right]. 
\ee
The first term in Eq. \ref{eq:FP} comes from the geometric path difference as a result of
the deflection, and the second term is the gravitational delay
described by the lens potential $\psi(\bmath{\theta})$.

The lens potential, $\psi(\bmath{\theta})$, is related to the projected dimensionless
surface mass density, $\kappa(\bmath{\theta})$, and the physical projected surface mass density,
$\Sigma(D_{\rm d} \bmath{\theta})$, by 
\be \label{eq:psiKappaDiffRelan}
\kappa(\bmath{\theta})= \frac {\Sigma(D_{\rm d} \bmath{\theta})} {\Sigma_{\rm cr}}=\frac{1}{2}\nabla^2\psi(\bmath{\theta}),
\ee
where 
\be \label{eq:sigcrit} 
\Sigma_{\rm cr} \equiv \frac{c^2 D_{\rm s}}{4 \pi G D_{\rm d} D_{\rm ds}}.
\ee
Equation \ref{eq:Tsimp} can be rephrased as
\be
\Delta t(\bmath{\theta},\bmath{\beta})  =  \frac{\Dt}{c},
\ee 
where $\Dt \equiv (1+\zd) D_{\rm d} D_{\rm s}/D_{\rm ds}$ is the time-delay distance.
Angular diameter distances are given by
\begin{equation}
D_{\mathrm{ij}}~=~ {c/H_0 \over {(1+z_{\mathrm{j}})}}\left( {\mathrm{sinn}\!\left( \sqrt{|\Omega_k|} \int_{z_{\mathrm{i}}}^{z_{\mathrm{j}}} {\mathrm{d} z \over E(z)}\right)\over \sqrt{|\Omega_k|}}\right)
\label{eq:Da}
\end{equation}
where $\mathrm{sinn}(x)~=~\sin(x)$, $x$, or $\sinh(x)$ for open, flat, or closed universes respectively, and $E(z)$ is the normalised Hubble parameter:
\begin{subequations}\label{Eq:Ez}
\begin{align}
E(z) &\equiv {H(z) \over H_0}\\
E^{\Lambda{CDM}}&~=~\!\sqrt{{\Omega_{\mathrm{M}}(1\!+\!z)^{3}\!+\!\Omega_k(1\!+\!z)^2\!+\!\Omega_{\Lambda}}}
\end{align}
\end{subequations}
Since angular diameter distances are proportional to the inverse of the Hubble constant, $1/H_0$,
so is $\Dt$ proportional to $1/H_0$, hence a measurement of $\Dt$ can be used to infer $H_0$
and also the other parameters in Equations \ref{eq:Da} and \ref{eq:Ez}.

\subsection{Mass-sheet degeneracy}
\label{sec:H0theory:MSD}

A dificulty with using time delay distances to contrain $H_0$ is the so called
Mass-sheet degeneracy \citep[see e.g.][for details]{Falco1985,Schneider2006}.
This is a degeneracy in the strong lens modeling corresponding to the addition of a
mass sheet that contributes a convergence and zero shear which combined with a
re-scaling of the original mass distribution leaves the predicted image positions and
magnifications unchanged. Given a lens
model $\kappa_{\rm lens}(\bmath{\theta})$, an alternative model described by the
transformation 
\be\label{eq:transform}
\widetilde{\kappa}_{\rm lens}(\bmath{\theta}) = \lambda + (1-\lambda)\kappa_{\rm lens}(\bmath{\theta}),
\ee
where $\lambda$ is a constant,would fit the image positions, shapes and relative fluxes equally well. Physically $\lambda$
is the \flag{efffective?} convergence of the mass-sheet and can take any real value. The mass-sheet degeneracy
corresponds to a degeneracy between this external convergence ($\kappax$) and the mass
normalization of the lens galaxy.

\flag{How do kappas combine? this is what Sherry said in suyuetal2010 - To be specific, the prescription that we adopt for
  combining the effects of many mass sheets at redshifts $z_i$ with
  surface mass densities $\Sigma_i$ is $\kappax=\frac{4\pi
    G}{c^2}\displaystyle\sum_i\frac{\Sigma_i(D_i\vec\theta)D_i D_{i{\rm s}}}{D_{\rm s}}$.}

Since strong lenses sit in our real Universe, they are not truely isolated;
there is some mass along the line of sight coming from intervening structures.
These structures may be physically associated with the lens e.g. other members
of a group or cluster, or they may be the result of chance alignments at any redshift. 
Whilst it is rare for three galaxies to line up well enough for both of the background sources
to be strongly lensed \citep{gavazzi2008,collett2012}, the shear scale of dark matter halos
makes a partial allignment -- with the additional structure acting as a weak lens -- 
a near certainty, indeed \citet{valewhite2003} and \citet{hilbert2007} find that there
are no empty lightcones in a realistic universe. It is necessary to include $\kappax$
in our lens modelling, if cosmological parameters are to be estimated accurately and precisely \citep{suyu2010}.

Despite the invariance of the image positions, shapes and relative fluxes
under the transformation in Eq. \ref{eq:transform}, the relative Fermat potential between the
images is not invariant, but changes according to
\be\label{eq:phitransform}
\Delta \widetilde{\phi_{\rm }}(\bmath{\theta},\widetilde{\bmath{\beta}_{\rm }}) 
= 
(1-\lambda) \Delta \phi_{\rm }
(\bmath{\theta},\bmath{\beta}_{\rm }).
\ee
If the source brightness changes the observed brightness in each image does not change
simultaneously - there is a time delay, $\Delta t$, between the brightness change in
each image. These time delays are inversely proportional to $H_0$ and
proportional to the relative Fermat potential (Equation \ref{eq:Tsimp}). Thus if
there is any external convergence, $\kappax$, that is not included in the lens
modeling, then the value of $H_0$ --inferred assuming $\kappax = 0$ -- will be
$1-\kappax$ less than the true value of $H_0$:
\be 
\label{eq:MassSheet:H0bias}
H_0^{\rm{true}}=(1-\kappax) H_0^{{\kappax = 0}}.
\ee
This degeneracy is present because lensing observations only deliver
relative positions and fluxes. However we can overcome this degeneracy 
if we have knowledge of $\kappax$.

For mass-sheets that are physically
associated with the lens galaxy, the mass sheet affects the stellar dynamics of the
lens galaxy. Stellar dynamics can be used jointly with lensing to break the internal
mass-sheet degeneracy by providing an additional estimate of the lens galaxy's mass
\citep[e.g.,][]{GroginNarayan96a, GroginNarayan96b, TonryFranx99, KoopmansTreu02,
 TreuKoopmans02, BarnabeKoopmans07, suyu2010}.

In this work our aim is to quantify the lensing contribution from mass sheets that
are not physically associated with the lens galaxy -- mass-sheets lying infront or behind
the lens \citep{FassnachtEtal06}.

@@

\section{Halo Model and Mass distribution}
\label{sec:model}

Weak lensing by NFW halos. Accounting for mass along the line of sight.
Approximations to full ray tracing: combining kappas: simple sum, Keeton
scaling, comments. Example ray traced kappa map from Stefan.
